$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-04-29 paper/paper.tex
Use the pi-calculus for the example of de Bruijn index usage

\paragraph*{Syntax}
We type the syntax for the \picalc{} in \S \ref{syntax}.
We use de Bruijn indices \cite{} as variable references, and dependent types to make the syntax well-scoped by construction: every unbound variable is accounted for in the type of the process that uses it.
In the \lambdacalc{} all computations are local and happen on syntactically contiguous terms; in the \picalc{} however communicating subprocesses can be non-contiguous and even have different parent processes.
One way to define a reduction relation modulo these syntactic minutiae is to define a structural congruence \cite{} relation that acts as a quotient type.
A term $A$ can then be freely rewritten into a term $B$ as long as $A$ is structurally congruent to $B$.
We type this equivalence relation in \S \ref{structural-congruence}.

@@

Data type definitions have double lines and index-free synonyms as rule names, for ease of reference.
Otherwise, we use the constructor name as the name of a typing rule.
We omit universe levels for brevity.
Implicit arguments are mentioned by type definitions but omitted in constructors.
$\type{TYPES}$ are in blue and uppercased (with indices as subscripts), $\constr{constructors}$ are in olive, $\field{fields}$ in red, and variables and functions in black.
Some constructors can be shared across types and disambiguated by context.
\todo{OD: the next part to be filled should be the notation, so that when I go through the paper, I can use this as a manual.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Syntax}
\label{syntax}

In the \picalc{} both scope restriction and input introduce abstractions.
The bodies of these constructs need a way of referring to their argument.
Using names as variable references is a popular option amongst humans.
However, names are cumbersome to mechanize: inserting a new variable into a context means proving that the name of such variable is different from all other variable names in context.
Moreover, names are of no significance to machines.

Machines need constructs they can algorithmically act upon.
The idea de Bruijn had \cite{} was to use a natural number $n$ (aka \emph{index}) to refer to the variable introduced $n$ binders ago.
Binders no longer introduce any names.
\todo{OD: IMPORTATNT!
There is too much on lambda-calculus; I suggest you give examples and illustrate your points on the pi-calculus only from now on. It can get too broad and lack focus otherwise... It is fine in the Introduction as you are giving an overview.}
\begin{example}
The expression $\lambda g . (\lambda f . f g) g$ in the \lambdacalc{} would translate to $\lambda (\lambda 0 1) 0$.
\end{example}
That is, terms at different \emph{depths} use different indices to refer to the same binding.
Humans often find this confusing.
There is however no reason not to annotate these indices with the original names bestowed by humans.
Machines can then manipulate references mechanically while still using names to present terms to humans.

>>>>

\paragraph*{Syntax}
We type the syntax for the \picalc{} in \S \ref{syntax}.
We use de Bruijn indices \cite{} as variable references, and dependent types to make the syntax well-scoped by construction: every free variable is accounted for in the type of the process that uses it.
In the \lambdacalc{} all computations are local and happen on syntactically contiguous terms; in the \picalc{} however communicating subprocesses can be non-contiguous and even have different parent processes.
One way to define a reduction relation modulo these syntactic minutiae is to define a structural congruence \cite{} relation that acts as a quotient type.
A term $A$ can then be freely rewritten into a term $B$ as long as $A$ is structurally congruent to $B$.
We type this equivalence relation in \S \ref{structural-congruence}.

@@

Data type definitions have double lines and index-free synonyms as rule names (for ease of reference).
We otherwise use the constructor name as the name of a typing rule.
Universe levels and universe polymorphism are omitted for brevity --- all our types are of type $\Set$.
Implicit arguments are mentioned by type definitions but omitted in constructors.
$\type{TYPES}$ are in blue and uppercased (with indices as subscripts), $\constr{constructors}$ are in olive, $\field{fields}$ in red, and variables and functions in black.
Some constructor names are overloaded --- they and disambiguated by context.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Syntax}
\label{syntax}

In the \picalc{} both scope restriction and input introduce abstractions.
The bodies of these constructs need a way of referring to their argument.
Using names as variable references is a popular option amongst humans.
However, names are cumbersome to mechanize: inserting a new variable into a context means proving that the name of such variable is different to all other variable names in context.
Moreover, names are of no significance to machines.

The idea de Bruijn had \cite{} was to use a natural number $n$ (aka \emph{index}) to refer to the variable introduced $n$ binders ago --- and binders no longer introduce names.
\begin{example}
The \picalc{} term $(\textbf{new} \; x) (x ( y ) . y \langle \texttt{"Hello world!"} \rangle . \mathbf{0} \parallel (\textbf{new} \; y) (x \langle y \rangle . y ( z ) .\mathbf{0}))$ would translate to $(\textbf{new}) (0 () . 0 \langle \texttt{"Hello world"} \rangle . \mathbf{0} \parallel (\textbf{new}) (1 \langle 0 \rangle . 0 () . \mathbf{0}))$.
\end{example}
That is, terms at different \emph{depths} use different indices to refer to the same binding.
Humans often find this confusing.
There is however no reason not to annotate these indices with the original names bestowed by humans.
Machines can then manipulate references mechanically while still using names to present terms to humans.
